Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-06-04 main.tex
Finished modifying paper to fit new result

Our discussion and experiments will focus on the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors, and a PCI card with a single high-end GPU.  
Recognition involving more than a few hundred users with contemporary hardware
would require a server(or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.\footnote{Non-shared memory configurations
include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.  
% WORKING HERE

@@

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  Therefore, if eight instances of
\eqref{eq:alignment-linearization} were be executed in parallel on the eight
cores, there would be an insufficient amount of L1 or L2 cache to hold $B_j$ and
$B_j^\dagger$ during the iterative process to execute Algorithm
\ref{alg:iterative_alignment}. Therefore, the bottleneck in this strategy becomes the
bandwidth between the L2 cache and L3 cache. The implementation of some
previous works \cite{WagnerA2009-CVPR,WagnerA2011-PAMI} is indeed based on this
mapping strategy.

In this paper, we show that for CPU parallelism of face alignment, a higher
performance implementation can be achieved by {\em not} taking advantage of
problem-level parallelism at all, and instead mapping the pixel-level parallelism
onto both the core-level and the SSE-level concurrency provided by the CPU.  In
other words, the proposed CPU implementation performs alignment of a query
image against the training images of a single user at a time using all cores on
both processors.

More specifically, in our implementation, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multi-core CPUs, and are able to
automatically utilize both core-level and vector-level concurrency. For other
operations that are not optimized by Intel in-house libraries, such as the shrinkage
operator in step 7 of Algorithm \ref{alg:iterative_alignment}, we achieve both
vector-level and core-level concurrency through a combination of the Open MP
API \cite{dagum2002openmp} and the automatic vectorization facilities of the
Intel C++ compiler (ICC) \cite{dulong1999overview}. 
%We shall compare the performance of the two parallelism methods in Section \ref{sec:experiment}.

\subsection{GPU Implementation}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, on the GPU cache is relatively unimportant for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes it's ability to sustain a very
high bandwidth to DRAM.  A higher overall bandwidth can be achieved if 
many memory transactions are performed simultaneously, than can be achieved
with a large number of smaller transactions that are just large enough to fill
the L2 cache.

Therefore, whereas on the CPU we advocate performing alignment for one training
class at a time, on the GPU we advocate performing alignment for many classes
concurrently.  \footnote{In CUDA terminology, Our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
actually get aligned concurrently is chosen by the GPU hardware. In our
experiments the hardware is able to schedule 4-5 classes per SMP to run
concurrently. This mapping of the problem-level parallelism onto the SMP-level
concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.

\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
\caption{Proposed mapping of alignment parallelism onto GPU concurrency}
\label{fig:alignment_mapping_gpu}
\end{figure}

The full resolution query image, along with its numerical derivatives in both
image directions, are stored in GPU texture memory. This  allows the
computation of $b(\tau)$ and $J(\tau)$, which essentially consist of warped
versions of the test image and its derivatives, to exploit the GPU's special
purpose hardware for bilinear interpolation.

Solving one alignment problem per SMP necessitated the implementation of a
custom CUDA routine for computing $B_j^\dagger = (B_j^TB_j)^{-1} B^T = G^{-1} B_j^T$. $G$
is inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Additionally, reductions on long vectors (i.e. sum or $\ell_1$ norm
of an image) are performed using a double precision temporary variable for accumulation.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

@@

\subsection{Recognition Stage Implementation} Compared to problem-level
parallelism, pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively easy to exploit using existing BLAS libraries.  For this reason,
both the CPU and GPU implementation of the recognition stage is very similar to
the CPU implementation of the $\ell_1$-min advocated in Section
\ref{sec:alignment_implementation_cpu}.  On the CPU, most of the operations map
well onto MKL BLAS calls, and the operations that do not can be easily
parallelized using OpenMP and auto-vectorization.

@@

\section{Experiments} \label{sec:experiment} In this section, we benchmark the
performance of our parallel implementations of $\ell_1$-min on CPU and GPU
platforms.  In order to show how our $\ell_1$-min algorithms scale with problem
size, in Section \ref{sec:simulation} we will begin with benchmarks for the
general $\ell_1$-min problem on synthetic data.  We will then progress in
Section \ref{sec:benchmark} to demonstrating the speed and accuracy of our
implementations as applied to the alignment and recognition stages of the face
recognition problem.

To generate a meaningful comparison between different hardware architectures,
in all cases we compare the performance of hardware architectures on a
per-board basis:  For CPU implementations, the benchmark makes use of all of
the cores in as many CPU's are present.  For GPU implementations, the benchmark
makes the best use of the entire GPU chip (most GPU boards have a single GPU
chip).  Aside from the exceptions already mentioned in Section
\ref{sec:alignment_implementation_gpu}, all math operations are performed in
single precision.

@@

The first experiment compares the performance of our two proposed
implementations of the general $\ell_1$-min (Algorithm \eqref{alg:alm}) given a
generic random matrix $A$ and a random sparse vector $\xx_0$.  The measurement
$d \times n$ matrix $A$ is generated as a random Gaussian matrix, with each
entry randomly generated from the standard normal and normalized with unit
column norm.  The ratio of $d/n$ is fixed at $1/2$ with $n$ varying from 1000 to 8000.
The ground truth signals, $\xx_0$ have a sparsity rate of 10\% of d with
elements sampled from a normal distribution and normalized to unit $\ell_2$ vector
norm.  Because the ground truth signal $\xx_0$ is known, the algorithm
terminats when $\|\bx-\bx_0\| < \tau$ with $\tau=10^{-3}$.  The measurement
vector is generated with the equation $\bb_0 = A \xx_0$.

The results of this benchmark can be found in Figure \ref{fig:random_data}.
The x-axis represents the size of the $A$ matrix and the y-axis represents the
average amount of time it to complete a single problem instance.  The GPU
implementation tends to be faster than the CPU implementation at solving a
single large problem, whereas the CPU implementation is faster at solving a
single small problem.  
\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/time_vs_matrix_size_constant_tol}
\end{center}
\caption{Comparison of $\ell_1$-min runtime vs. dimensions of $A$ on random data.}
\label{fig:random_data}
\end{figure}

@@

Figure \ref{fig:alignment_stage_runtime} shows the average run time of the CPU and GPU implementations to align
a query image against all the subject classes separately. We vary the number of
subject classes to show how the algorithms scale. 
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_runtime_graph}
\caption{Alignment stage runtime vs. size of training database.}
\label{fig:alignment_stage_runtime}
\end{figure}
Similar to the results in Figure \ref{fig:random_data},
there is a crossover point above which the GPU out-performs the CPU.  Since
the CPU algorithm solves problems sequentially, there is a linear trend.  
The tiers seen in the GPU profile result from the GPU hardware scheduling alignment
problems in concurrent batches, but the overall trend is linear as expected.

@@

\section{Conclusion}
We have shown that our proposed parallelization is {\em eight times} faster
than an implementation solving a single problem at a time using the stock
CUBLAS library. Furthermore, the proposed mapping of the parallelism of the
face alignment problem onto the concurrency provided in modern CPU and GPU
architectures results in dramatic improvements in recognition speed over
previously reported implementations of the same pipeline.  Whereas earlier
implementations required over a second to align a test image against just one
of the 250 users in Multi-PIE session 1, the proposed CPU implementation can
perform the same task in only 45 seconds, and the GPU implementation can do it
in {\bf under 10 seconds}!.  In principle, the parallelization techniques
presented here make it feasible to implement an access control system for
hundreds of users on a handful of purpose-build multi-GPU workstations.

>>>>

Our discussion and experiments will address the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors, and a PCI card with a single high-end GPU.  
Recognition involving more than a few hundred users with contemporary hardware
would require a server(or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.\footnote{Non-shared memory configurations
include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.  
% WORKING HERE

@@

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  
In this paper we consider two mappings of the parallelism available in the alignment
stage to the concurrency available on the CPU.  We argue that the latter
implementation should achieve higher performance, and later verify that this is the case.

The first implementation disregards 
problem-level parallelism, and instead mapping the pixel-level parallelism
onto both the core-level and the SSE-level concurrency provided by the CPU.  In
other words, alignment is performed of the query
image against the training images of a single user at a time using all cores on
both processors.  The advantage of this implementation is that all of the variables
for the inner loop, most notably $B_j$ and $B_j^\dagger$, fit in L2 cache.

The second implementation maps problem-level parallelism onto core-level
concurrency, and maps pixel-level parallelism onto SSE-level concurrency.  In
other words, eight instances of \eqref{eq:alignment-linearization} are executed
in parallel with each problem running on one of the eight cores.  The advantage
of this implementation is that since the cores are operating on different
alignment problem instances, no data is shared between cores, and therefore no
synchronization is necessary between the cores.  Furthermore, even with eight
problem instances running concurrently, the local variables for the inner
loop still fit in L3 cache.  Thanks to the sequential data access patterns of
the solver, prefetching significantly reduces the bandwidth difference between
L2 and L3 cache.  For these reasons, we expect this implementation to out-perform
the previous implementation.

%The implementation of some
%previous works \cite{WagnerA2009-CVPR,WagnerA2011-PAMI} are indeed based on the latter
%parallelization strategy.

%64 byte cache line-size per clock.  L3 cache running at lower clock than cpu.  Two reads and a write per clock.
%4 floats at a time in SSE.

In both implementations, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multi-core CPUs, and are able to
automatically utilize both core-level and vector-level concurrency (for the first implementation),
but are also available in single-threaded versions (for the second implementation). For 
operations that are not optimized by Intel in-house libraries, such as the shrinkage
operator in step 7 of Algorithm \ref{alg:iterative_alignment}, we achieve 
vector-level concurrency via
the automatic vectorization facilities of the Intel C++ compiler (ICC) \cite{dulong1999overview}.
To achieve core-level concurrency we make use of the Open MP API. \cite{dagum2002openmp} 
%We shall compare the performance of the two parallelism methods in Section \ref{sec:experiment}.

\subsection{GPU Implementation}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, on the GPU cache is relatively unimportant for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is it's ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing memory requests simultaneously.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

Therefore, our recommended GPU implementation is strikingly similar in spirit
to our recommended CPU implementation: on the CPU we use the cores to solve
multiple instances concurrently, and on the GPU we use multiple SMP's for the 
same purpose.
\footnote{In CUDA terminology, Our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
actually get aligned concurrently is ultimately chosen by the GPU hardware, 
but can be influenced by manual tuning of the code. We have empirically determined
that performance is highest with 5-7 subject classes aligned simultaneously on each SMP.
%This mapping of the problem-level parallelism onto the SMP-level
%concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.
%\begin{figure}
%\centering
%\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
%\caption{Proposed mapping of alignment parallelism onto GPU concurrency}
%\label{fig:alignment_mapping_gpu}
%\end{figure}

Several other aspects of the implementation merit discussion.  
The full resolution query image, along with its numerical derivatives in both
image directions, are stored in GPU texture memory. This  allows the
computation of $b(\tau)$ and $J(\tau)$, which essentially consist of warped
versions of the test image and its derivatives, to exploit the GPU's special
purpose hardware for bilinear interpolation.

Solving one alignment problem per SMP necessitates the implementation of a
custom CUDA routine for computing $B_j^\dagger = (B_j^TB_j)^{-1} B^T = G^{-1} B_j^T$. $G$
is inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Additionally, reductions on long vectors (i.e. sum or $\ell_1$ norm
of an image) are performed using a double precision temporary variable for accumulation.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

@@

\subsection{Recognition Stage Implementation} Compared to problem-level
parallelism, the exclusively pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively straightforward to exploit:  
On the CPU, most of the operations map
well onto MKL BLAS calls, and operations that do not can be easily
parallelized using OpenMP and auto-vectorization.

@@

\section{Experiments} \label{sec:experiment} In this section, we benchmark the
performance of our parallel implementations of $\ell_1$-min on CPU and GPU
platforms.  In order to show how our $\ell_1$-min algorithms scale with problem
size, in Section \ref{sec:simulation} we will begin with benchmarks for the
general $\ell_1$-min problem on synthetic data.  We will then progress in
Section \ref{sec:benchmark} to demonstrating the speed and accuracy of our
implementations as applied to the alignment and recognition stages of the face
recognition problem.
%To generate a meaningful comparison between different hardware architectures,
%in all cases we compare the performance of hardware architectures on a
%per-board basis:  For CPU implementations, the benchmark makes use of all of
%the cores in as many CPU's are present.  For GPU implementations, the benchmark
%makes the best use of the entire GPU chip (most GPU boards have a single GPU
%chip).  
Aside from the exceptions already mentioned in Section
\ref{sec:alignment_implementation_gpu}, all math operations are performed in
single precision.

@@

The first experiment compares the performance of our proposed CPU and GPU
implementations of the general $\ell_1$-min (Algorithm \eqref{alg:alm}), given a
generic random matrix $A$ and a random sparse vector $\xx_0$.  The measurement
$d \times n$ matrix $A$ is generated as a random Gaussian matrix, with each
entry randomly generated from the standard normal and normalized with unit
column norm.  The ratio of $d/n$ is fixed at $1/2$ with $n$ varying from 1000 to 8000.
The ground truth signals, $\xx_0$ have a sparsity rate of 10\% of d with
elements sampled from a normal distribution and normalized to unit $\ell_2$ vector
norm.  Because the ground truth signal $\xx_0$ is known, the algorithm
terminates when $\|\bx-\bx_0\| < \tau$ with $\tau=10^{-3}$.  The measurement
vector is generated with the equation $\bb_0 = A \xx_0$.

The results of this benchmark can be found in Figure \ref{fig:random_data}.
The x-axis represents the size of the $A$ matrix and the y-axis represents the
average amount of time it to complete a single problem instance.  The GPU
implementation tends to be faster than the CPU implementation at solving a
single large problem, whereas the CPU implementation is faster at solving a
single small problem.  The transition between the two regimes occurs where
the problem size begins to exceed the size of the CPU L3 cache. 
\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/time_vs_matrix_size_constant_tol}
\end{center}
\caption{Comparison of $\ell_1$-min runtime vs. dimensions of $A$ on random data.}
\label{fig:random_data}
\end{figure}

@@

Figure \ref{fig:alignment_stage_runtime} shows the average run time of the CPU and GPU implementations to align
a query image against all the subject classes separately. We vary the number of
subject classes to show how the algorithms scale. 
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_runtime_graph}
\caption{Alignment stage runtime vs. size of training database.}
\label{fig:alignment_stage_runtime}
\end{figure}
The tiers seen in the GPU profile result from the GPU hardware scheduling alignment
problems in concurrent batches, but the overall trend is linear as expected.
The CPU implementation that solves one problem instance per core out-performs
the single-instance-at-a-time implementation by a wide margin, and is even competitive
with the GPU implementation.\footnote{Note that the GPU implementation was tuned
to be fast for the full problem set; tuning affects the locations of the tiers,
so the lower corners of the tiers are most relevant.}

@@

\section{Conclusion}
We have demonstrated that on both CPU and GPU algorithms, parallelizations of
ALM that solve multiple face alignment problems concurrently are significantly
faster than implementations that rely purely on vendor-supplied BLAS libraries.
Furthermore, thanks to a combination of faster hardware, and more efficient use
of the hardware, we have demonstrated dramatic improvements in recognition
speed over previously reported implementations of the same pipeline.  Whereas
earlier implementations required over a second to align a test image against
just one of the 250 users in Multi-PIE session 1, the proposed 
implementations can perform the same task in 
{\bf under 10 seconds}!.  In principle, the
parallelization techniques presented here make it feasible to implement an
access control system for hundreds of users on a handful of purpose-build
multi-GPU workstations.  As CPU manufacturers 
increase the number of cores and vector widths, and as GPU manufacturers increase the amount
of cache, both architectures are rapidly converging towards an architectural balance
that increasingly favorable for $ell_1$-min based face recognition.
